---
title: "Preliminary Report"
author: "Derwin McGeary"
date: "20 December 2015"
output: html_document
---
## Introduction
This is a preliminary report on the datasets for the Capstone Project. I will give a quick summary of the contents and nature of the data.

## The basics

```{r, echo=FALSE, cache=TRUE}
library("tm")
library("RWeka")
library("ggplot2")
news <- readLines("final/en_US/en_US.news.txt")
blogs <- readLines("final/en_US/en_US.blogs.txt")
tweets <- readLines("final/en_US/en_US.twitter.txt")
```

We have three large files of English text, from three sources: News articles, blogs, and tweets. Each line contains one article, blog or tweet. We have a fairly large amount of data, `r length(tweets)` tweets, `r length(blogs)` blogs, and `r length(news)` news articles.

### The Tweets
```{r, echo=FALSE, cache=TRUE}
tlengths <- nchar(tweets)
tsummary <- summary(tlengths)
ggplot(as.data.frame(tlengths), aes(x=tlengths)) + ggtitle("Distribution of Tweet Lengths") + 
    geom_histogram(binwidth=1, colour="white", fill="#55acee") + theme_bw()
```

This plot shows that the longest tweet is `r max(tlengths)` characters, and it we can see that there is a hard limit as that is the most popular length, although there is a wide distribution of lengths.


### The News
```{r, echo=FALSE, cache=TRUE}
nlengths <- nchar(news)
nsummary <- summary(nlengths)
ggplot(as.data.frame(nlengths), aes(x=nlengths)) + ggtitle("Distribution of News Article Lengths") + 
    geom_histogram(binwidth=10, colour="white", fill="#BB1919") + theme_bw()
```

### The Blogs

```{r, echo=FALSE, cache=TRUE}
blengths <- nchar(blogs)
bsummary <- summary(blengths)
ggplot(as.data.frame(blengths), aes(x=blengths)) + ggtitle("Distribution of Blog Post Lengths") + 
    geom_histogram(binwidth=1, colour="black", fill="white") + theme_bw()
```

